---
output: github_document
---

<!-- README.md is generated from README.Rmd. -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.path = "man/figures/README-",
  message=FALSE
)
```

[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![Build Status](https://travis-ci.org/Beirnaert/MetaboLouise.svg?branch=master)](https://travis-ci.org/Beirnaert/MetaboLouise)

METABOlomics LOngitudinal SimUlation Engine. A small package for simulating longitudinal/dynamic metabolomics data based on underlying biological networks. 

## Overview

MetaboLouise has been built to simulate dynamic metabolomics datasets. However, this focus on metabolites is not necessary. The problem can just as easily be pictured as nodes (metabolites) in a network whereby the values of the nodes (concentrations) evolve over time depending on the topology of the network and the rates (enzyme concentrations) assigned to the edges. 

The basics are explained below and more specific information can be found in the vignette (on [CRAN](https://cran.r-project.org/package=MetaboLouise)). 


## Installation

```{r, eval = FALSE}
install.packages("MetaboLouise")

# Or the the development version from GitHub (requires devtools package):
# install.packages("devtools")
devtools::install_github("Beirnaert/MetaboLouise")
```


## Usage

The proof of the pudding is in the eating. Hence, the most basic example.


```{r example, dpi=400, fig.width=6, fig.height=4}
#library(MetaboLouise)
devtools::load_all()
set.seed(1)
Nmetabos <- 20
Nrates <- 10
 
Network <- NetworkCreateR(N = Nmetabos, plothist = FALSE)
 
     
Example_simulation <- DataSimulateR(NetworkMatrix = Network, 
                                    dT = 0.01, 
                                    Tstart = 0, 
                                    Tstop = 3,
                                    T0_nodes = 100)

```

